From correlation to causality: statistical approaches to learning regulatory relationships in large-scale biomolecular investigations

### 2015-12-18 manuscript/write_up.tex
a bit late in the game

\added{To understand why, recall that we are interested in paring down the correlation graph in step one to the sparse conditional independence graph in step 2.  A {\it true positive} is the case when we correctly keep an edge from step 1 to step 2 in cases where the proteins are not conditionally independent.  Incorrectly keeping an edge when the proteins are actually conditionally independent is a case of a false positive.  Since the edges we preserve imply a regulatory relationship and motivate the expense of a validation experiment, false positives are costly and should be minimized.  However, the number of false positives increases as an exponential rate as one increases the number of proteins, much faster than the number of true positives.}

In Step 1, the challenge is in quantifying statistical associations between each pair of the analytes across the biological samples. A large number of analytes yields a large number of spurious statistical associations, which arise without any biological justification, and are purely an artifact of random chance. Systematic pairwise relationships such as between Raf, Mek and Erk in the MAPK pathway will be obscured by the many spurious relationships that they will each form with causally unrelated proteins.

We illustrate this problem with a computer simulation, inspired by Fan et. al \cite{fan2014challenges} but translated to our context.  We simulate data from a conditional independence network of 60 proteins and evaluate performance of structure inference algorithms in recovering this structure.  We then simulate a second experiment where the number of proteins is increased 99, to show how the increased protein coverage affects the performance of structure inference.  We varied depth as well as breadth -- In order to evaluate the effect of increasing sample size we start with 100 replicates, then repeat the two experiments with 200 replicates.  We evaluate performance by counting the total number of false positives (reject conditional independence for two conditionally independent proteins) and true positives (correct rejections).

The simplest case of for testing conditional independence is one of three variables -- two to compare, and at one variable upon which to condition the comparison.  Thus in the simulate we generate networks composed of three-protein subnetworks comprised of protein A, B, and C.  C's measurements come a univariate Gaussian distribution with 0 mean and variance 1.  Proteins A and B also have 0 mean and variance of 1, but their dependence with C is quantified in the multivariate Gaussian covariance matrix, each has a covariance with C of .3.  Most importantly, A and B are conditionally independent of each other given C.  So each subnetwork has two edges, from A to C and from B to C, and no edge from A to B is missing.  We created a network of 60 proteins within 20 disjoint subnetworks, 40 edges in total, and simulated multivariate Gaussian data with the covariance structure determined by this network.

We examined two cases for expanding the number of measured proteins.  The first describes the case of measuring proteins that provide no additional information about the biology.  We model this scenario by adding 39 unconnected proteins with independent Gaussian values for the second experiment.  Any inferred edge in the condition independence network linking these two proteins would be a false positive.  We then look at the case where more proteins are added, but the added proteins are informative about the biology.  We model this by increasing the number of triplets from 20 to 33, again to 99 proteins in total.

\begin{figure}[!tpb]
\centerline{\includegraphics[width=.8\textwidth]{figs/spurious_corr2.png}}
\caption{Simulated experiments where at first 60 proteins are quantified, after which throughput is increased by an additional 2940 proteins.  This scenario is simulated 500 times an each time the maximum Pearson correlation across any pair of proteins is collected. The histograms display the distributions of the maximum correlation values.Parts A and B indicates the case of randomly adding completely independent proteins to the panel, parts C and D illustrated informed addition of proteins -- the added proteins contain interdependencies.  The left column (A and C) show results for 30 replicates, the right (B and D) for 60 replicates. Comparing A and B to C and D demonstrate increasing the number of proteins on the panel creates more spurious correlation, and this is unaffected by whether or not the addition of proteins are added in an informative way.  This is a problem because correlation is used as initial evidence for a biological relationship.  Comparing the left and right columns show increasing the number of replicates can mitigates this somewhat.}
\label{spur_corr}
\end{figure}

Figure \ref{spur_corr} shows illustrates simulated experiments where at first 60 proteins are quantified, after which throughput is increased by an additional 2940 proteins.  This scenario is simulated 500 times an each time the maximum Pearson correlation across any pair of proteins is collected. The histograms display the distributions of the maximum correlation values.  The top row (figure \ref{spur_corr} parts A and B) indicates the case if randomly adding proteins to the panel -- the added proteins where completely independent of one another protein in the network.  The bottom row (C and D) indicates the case of informed addition of proteins to the panel -- the added proteins contain interdependencies.  The first column (A and C) show results for 30 replicates, the second (B and D) for 60 replicates.  Parts A and B show that an increase in proteins leads to more high correlations, however since the added proteins were completely independent, the higher correlations were completely spurious.  Further, comparing parts C and D to A and B show that even when addition of proteins is informed by biology, the results are the same.  Spurious correlation increases with the number of proteins whether or not the addition of proteins is informed by biological prior knowledge.  This is clearly a problem when high Pearson correlation is used as an initial evidence of a biological function.  Comparing the left and right columns show that increasing the number of replicates can mitigate this situation someone, but as we will see next, false positives will remain a problem.

This increased incidence of spurious correlations impedes Step 2, which \added{infers} conditional independences in the data.  To illustrate this, we repeated the previous simulation, but this time in each iteration we \added{performed Step 1 by calculating pairwise \added{Pearson} correlations, and then Step 2 by applying the Grow-Shrink algorithm \cite{margaritis2003learning} for causal inference, an instance from a class of algorithms called constraint-based network inference algorithms\cite{spirtes2000causation} that proceed with a series of conditional independence tests that control the false positive rate.}   \added{These algorithms have stricter data requirements that correlation calculations, so we changed settings to 60/100 samples, and 60/99 proteins to meet these requirements.}

Figure \ref{spur_dep} shows the counts of false positive edges in the resulting conditional independence graph, i.e. the number of protein pairs \added{where the null hypothesis of conditional independence was unduly rejected}.  The results demonstrate that an increase in the number of proteins leads to an increase in false positive rejections of conditional independence, and to a denser conditional independence graph.  \added{Again, comparing the top and bottom rows in figure \ref{spur_dep} show that it doesn't matter if the addition of proteins to the panel is informed or not. Increasing the number of proteins results in an exponential increase in pairwise comparisons -- a 65\% increase from 60 to 99 proteins results in a 175\% increase in the number of pairs.  So though the algorithm controls the rate of false positives, the absolute number of false positives increases exponentially.  Increasing the number of replicates does not help -- paradoxically in this case it hurt, most likely because the additional degrees of freedom enabled more tests, some of which added to the false positive count.} In other words, the computational methods for causal inference are more likely to fail in a large-scale experiment, because they cannot reliably distinguish the potential causal relationships from noise.  \added{ To verify that this pattern is not the artifact of a particular protein inference algorithm we repeated the simulation with two comparable algorithms\cite{tsamardinos2003algorithms,yaramakala2005speculative}, and arrived to the same conclusion (results not shown).}

\begin{figure}[!tpb]
\centerline{\includegraphics[width=.8\textwidth]{figs/spurious_dep.png}}
\caption{Simulated experiments quantifying 60 and 99 unrelated proteins in 60 and 100 biological samples. The histograms display the counts of false positives in conditional independence tests between pairs of proteins, calculated over 500 repetitions of the simulation. The top row parts A and B illustrate uninformed addition of 39 proteins to the starting amount of 60, the bottom parts C and D illustrate informed addition.  The left column parts A and C show results with 60 replicates, the right column B and D for 100 replicates.  Increasing the number of proteins results in more false positives without any biological justification.  Informed addition and added replicates does not help.}
\label{spur_dep}
\end{figure}

\added{These results lead to obvious question.  What exactly is the use of having more replicates? How does an informed selection of proteins for the panel make a difference relative to random selection?  These factors come to play in detection of true positives.  Figure\ref{true positives} is the true positive counterpart to figure\ref{true_pos}.  Given the described subnetwork approach we described above, the 60 protein network is comprised of 20 subnetworks of 3 proteins, each with 2 edges each for a total of 40 relationships that the algorithm is tasked with detecting.  That upper bound is illustrated with the dashed line in figure\ref{true_pos} parts A and B.  Uninformative addition of proteins adds no new relationships.  In parts C and D, informed addition of proteins adds 26 new relationships, indicated by the double-dashed line.  Comparing parts A and B show that increasing the number of replicates increases the amount of true positives.  Parts C and D show that when there are more total relationships to detect, having more replicates is essential -- when more relationships were added without increasing replicates, the count of true positives actually decreased slightly on average as the algorithm tries to do more with little.  More importantly, the scale of the increase in total true positives gained from informed addition of proteins to the panel was less than that of the increase of false positives.  The costs of spurious correlation and resulting false positives quickly outweighs the benefits of higher throughput}

\begin{figure}[!tpb]
\centerline{\includegraphics[width=.8\textwidth]{figs/true_pos.png}}
\caption{Simulated experiments quantifying 60 and 99 unrelated proteins in 60 and 100 biological samples. The histograms display the counts of true positives in conditional independence tests between pairs of proteins, calculated over 500 repetitions of the simulation. The top row parts A and B illustrate uninformed addition of 39 proteins to the starting amount of 60, the bottom parts C and D illustrate informed addition.  The left column parts A and C show results with 60 replicates, the right column B and D for 100 replicates.  Dashed lines indicate the total amount of positives the algorithm is tasked with discovering.  Increasing the number of replicates increases true positives and is indeed essential when increasing the number of proteins increases the number of relationships to be detected.}
\label{spur_dep}
\end{figure}

>>>>

\added{To understand why, recall that we are interested in paring down the correlation graph in step 1 to the sparse conditional independence graph in step 2.}  In Step 1, the challenge is in quantifying statistical associations between each pair of the analytes across the biological samples. A large number of analytes yields a large number of spurious statistical associations, which arise without any biological justification, and are purely an artifact of random chance. Systematic pairwise relationships such as between Raf, Mek and Erk in the MAPK pathway will be obscured by the many spurious relationships that they will each form with causally unrelated proteins.

We illustrate this problem with a computer simulation, inspired by Fan et. al\cite{fan2014challenges} but translated to our context. First, we simulated an experiment that quantifies the abundances of 60 proteins in 60 biological samples. Second, we simulated another experiment where the number of proteins was increased from 60 to 6000. In both experiments the proteins are completely independent from each other, and each protein in each replicate is assigned a value randomly drawn from a Gaussian distribution. In other words, we do not expect any biologically meaningful associations in these data. We repeated each of these simulations 500 times. Figure \ref{spurious_corr.png} shows for each experiment the histogram of the highest Pearson correlation across any pair of proteins in the 500 instances of the simulation. As can be seen, the increase to 6000 proteins produces relatively large maximum pairwise correlations, demonstrating that an increase in the number of proteins leads to an increase in spurious correlation.  This is clearly a problem when high Pearson correlation is used as an initial evidence of biological function.

\begin{figure}[!tpb]
\centerline{\includegraphics[width=.8\textwidth]{figs/spurious_corr.png}}
\caption{Histogram of maximum observed correlation values between unrelated protein pairs in a simulation of 500 experiments.  In each iteration,  first 60 proteins are quantified, after which throughput is increased to 6000 proteins.  Increasing the number of proteins increases spurious correlation.  This is a problem because correlation is used as initial evidence for a biological relationship.}
\label{spur_corr}
\end{figure}

Similarly, the increased incidence of spurious correlations impedes the performance of statistical methods in Step 2, which elucidate conditional independences in the data.  If we think of this step as \added{attempting to detect which pairs are not conditionally independent}, then the spurious correlations result in more false positives \added{detections}. To illustrate, \added{we performed a second simulation}, this time we included some dependence between proteins so we could evaluate the effect of dimensionality of the data on true positives (pairs of proteins that are not conditionally independent) as well as false positives.

\added{To build dependence into the simulated data, we created multiple sets of 3-proteins "triplets" A, B, and C, where all three are dependent but A and B are conditionally independent given C.  Thus perform perfectly on the $i^{th}$ triplet, the algorithm would need two true positives (rejecting conditional independence between $A_i$ -- $C_i$ and $B_i$ -- $C_i$), find the $A_i$ and $B_i$ pair conditionally independent, and find any pair containing a node in the $i^th$ triplet and any node not in the $i^th$ triplet to be independent.  We made it such that the proteins in each triplet generate measurements with Gaussian values with mean 0 and variance 1, and that the covariance between A--C and B--C was .3.  We simulated data from a multivariate Gaussian. }

\added{Again we run the simulation 500 times, in each iteration we performed step 1 by calculating pairwise Pearson correlations, and then Step 2 by applying the Grow-Shrink algorithm \cite{margaritis2003learning} for causal inference, an instance from a class of algorithms called constraint-based network inference algorithms\cite{spirtes2000causation} that proceed with a series of conditional independence tests that control the false positive rate.  To assure we weren't observing algorithm specific effects, we also repeated the simulation other algorithms from this class\cite{tsamardinos2003algorithms}\cite{yaramakala2005speculative}.  The results did not differ (results not shown).}

\added{These algorithms have stricter data requirements that correlation calculations, so in this simulation we increase from 60 proteins to only 99.  The starting 60 proteins contains 20 triplets, and thus perfect performance of the algorithm would yield 40 true positives (and of course 0 false positives).  In a high-throughput experiment, the investigator may arbitrarily select proteins for quantification, hoping the more proteins are quantified, the higher the chance of uncovering biology.  Alternatively, he or she may apply their prior knowledge about the system to select proteins that are strong candidates for being involved in the biology.  We respectively term these approaches "uninformed addition" and "informed addition" of proteins.  We simulate "uniformed addition" but adding 39 proteins to the original 60 that are completely independent.  In this case perfect performance is still 40 true positives.  In the second "informed addition" case, we add 13 new triplets, increasing the perfect performance standard from 40 to 66.  In addition to comparing these two strategies, we compare the 60 replicate case to the 100 replicate case, to show what role sample size has in affecting performance.}

Figure \ref{spur_dep} shows the counts of false positive edges in the resulting conditional independence graph, i.e. the number of protein pairs \added{where the null hypothesis of conditional independence was unduly rejected}.  The results demonstrate that an increase in the number of proteins leads to an increase in false positive rejections of conditional independence, and to a denser conditional independence graph.  \added{This occurs because the number of false positives depends on the total number of pairwise comparison -- a 65\% increase from 60 to 99 proteins results in a 175\% increase in the number of comparisons.  In other words, the computational methods for causal inference are more likely to fail in a large-scale experiment, because they cannot reliably distinguish the potential causal relationships from noise.}

\begin{figure}[!tpb]
\centerline{\includegraphics[width=.8\textwidth]{figs/spurious_dep2.png}}
\caption{Histograms of false positives detection of relationships between proteins (rejections of tests for conditional independence) over 500 simulated experiment.  Simulated experiments start with quantifying 60 and then expand to 99.  The initial 60 proteins contains 40 pairwise relationships that the algorithm is based with detection.  In parts A and B, 39 proteins are added that have no relationship with any other protein.  In parts C and D, 39 proteins are added that have 26 additional pairwise relationships. The left column parts A and C show results with 60 replicates, the right column B and D for 100 replicates.  Increasing the number of proteins results in more false positives without any biological justification.  Informed addition of proteins mitigates this only slightly.  Adding replicates paradoxically increases false positives.  In both cases there are less replicates than pairwise comparisons, so the extra replicates add degrees of freedom which enables more tests, some of which add to the total false positives.}
\label{spur_dep}
\end{figure}

\added{Comparing the panels figure \ref{spur_dep} reveals a noteworthy result; informed addition of proteins offers only a slight reduction in false positives on average.  Increasing sample size doesn't help much either, paradoxically in this case it hurt, because even at 100 replicates there are still  more pairwise comparisons than replicates, so the algorithm uses every last degree of freedom at its disposal.  Increasing the number of replicates from 60 to 100 therefore allowed more conditional independence tests to take place, and some of these additional tests added to the false positive count.}

\begin{figure}[!tpb]
\centerline{\includegraphics[width=.8\textwidth]{figs/true_pos.png}}
\caption{Simulated experiments quantifying 60 and 99 unrelated proteins in 60 and 100 biological samples. The histograms display the counts of true positives in conditional independence tests between pairs of proteins, calculated over 500 repetitions of the simulation. The top row parts A and B illustrate uninformed addition of 39 proteins to the starting amount of 60, the bottom parts C and D illustrate informed addition.  The left column parts A and C show results with 60 replicates, the right column B and D for 100 replicates.  Dashed lines indicate the total amount of positives the algorithm is tasked with discovering.  Increasing the number of replicates increases true positives and is indeed essential when increasing the number of proteins increases the number of relationships to be detected.}
\label{true_pos}
\end{figure}

\added{So what exactly is the use of having more replicates? How does an informed selection of proteins for the panel make a difference relative to random selection?  These factors come to play in detection of true positives.  Figure \ref{true_pos} is the true positive counterpart to figure\ref{spur_dep}.  Parts A and B illustrate the 40 true positives upper bound on performance with the dashed line, and in parts C and D, where "informed addition" of proteins is applied, the increase upper bound of 66 is pictured with the double-dashed line.  Comparing parts A and B show that increasing the number of replicates increases the amount of true positives.  Parts C and D show that when there are more total relationships to detect, having more replicates is essential -- when more relationships were added without increasing replicates, the count of true positives actually decreased slightly on average as the algorithm tries to do more with little.  However, when comparing parts C and D in figure \ref{true_pos} to C and D in figure \ref{spur_dep}, we see that the scale of the increase in total true positives gained from informed addition of proteins was less than that of the increase of false positives.  The spurious correlation and resulting false positives in high throughput experiments quickly outweigh the benefit that informed selection of proteins and greater sample size has on discovery.}
